---
title: "Week 4 Exercises"
author: "J.Blondeau"
date: "2023-09-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This week we will continue wrangling data in the Tidyverse.  The following are some very common functions you will use in your everyday data world.

- merge()
- mutate()
- case_when()
- group_by()
- summarize()


All the exercises for this week use the "PSU_Fish_Density.csv" and the "taxonomic_data.csv" data sets.

```{r}
library(tidyverse)
density <- read.csv("data/PSU_Fish_Density.csv")
View(density)
tax <- read.csv("data/taxonomic_data.csv")
View(tax)
```
1. Add the common name (only common name) of each species in the density data set as a new field.  
    - Reorder the columns to be more readable.
    
```{r}
density1 <- left_join(density, tax, by = "SPECIES_CD")
View(density1)

density1 <- density1 %>% select (YEAR, REGION, STRAT, PROT, PRIMARY_SAMPLE_UNIT, SPECIES_CD, COMNAME, density)
View(density1)
```

2. Add a new field to the density data set named "fishery_target" at set all to TRUE

```{r}
density2 <- density %>% mutate(fishery_target = TRUE)
View(density2)
```

3. Add a new field named "group" at set the value based on fish species where...
    - grouper = Coney, Red hind, 
    - sanpper = Mutton, Gray, Yellowtail
    - parrotfish = Stoplight
    - other = Triggerfish, Hogfish
    
```{r}
density1 <- density1 %>% mutate(group = case_when(COMNAME %in% c("coney", "red hind")~"grouper", COMNAME %in% c("mutton snapper", "gray snapper", "yellowtail snapper")~"sanpper", COMNAME %in% c("stoplight parrotfish")~"parrotfish", COMNAME %in% c("triggerfish", "hogfish")~"other"))
View(density1)

```
4. Using the group_by function, how many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR?
```{r}
psu <- density %>% group_by(YEAR) %>% summarise(length(unique(PRIMARY_SAMPLE_UNIT)))

View(psu)

```

5. How many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR and PROT combination?

```{r}
yearprot <- density %>% group_by(YEAR, PROT) %>% summarise(length(unique(PRIMARY_SAMPLE_UNIT)))

View(yearprot)

```
6. How many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR, PROT and STRAT combination?

```{r}
yearprotstrat <- density %>% group_by(YEAR, PROT, STRAT) %>% summarise(length(unique(PRIMARY_SAMPLE_UNIT)))

View(yearprotstrat)

```

7. What is the difference between the following two expressions?  The summarize and mutate calls after the group_by do very different things.  In what situations would you use them?
    
    - data %>% # data here is the "PSU_Fish_Density" data set
        <br></br>
        group_by(YEAR) %>% 
        <br></br>
        summarize( n = n_distinct(PRIMARY_SAMPLE_UNIT))
        
Summarize calls will use data from the current dataset to modify/summarize. 

    - data %>%
        <br></br>
        group_by(YEAR) %>% 
        <br></br>
        mutate( n = n_distinct(PRIMARY_SAMPLE_UNIT))
        
Mutate calls will create a brand new column or change one in a dataframe. 
        
        
8. Create a new dataframe that shows mean density of each species per year...hint
    - YEAR  | SPECIES_CD  | meanDensity
      ----- | ----------  | -----------
      2017  | BAL VETU    |    0.224   
      2019  | CEP FULV    |    0.685
      
```{r}
mean <- density %>% group_by(YEAR, SPECIES_CD) %>% summarise(mean(density))
View(mean)

```
      
9. Create a new dataframe that shows mean density of each species in each PROT per year...hint
    - YEAR  | SPECIES_CD  | PROT | meanDensity
      ----- | ----------  | -----|-----------
      2017  | BAL VETU    | 0    |0.229  
      2019  | CEP FULV    | 1    |0.308
      
```{r}
mean1 <- density %>% group_by(YEAR, SPECIES_CD, PROT) %>% summarise(mean(density))
View(mean1)

```

10. In the "PSU_Fish_Density.csv" data set, the PROT field refers to sites that are inside the VI National Park (PROT = 1) and sites that are outside the Park (PROT = 0).  How many of the 8 species had higher densities inside the National Park?

2 species had higher densities. 

```{r}
mean3 <- density %>% group_by(SPECIES_CD, PROT) %>% summarise(mean(density))
View(mean3)

```





